The past and future of visualization for computing in science and engineering

### 2020-03-27 CombaWeiskopf2020/paper.tex
edits section 2 and fin

\begin{figure*}
    \begin{center}    
        \includegraphics[width=0.55\textwidth]{museum1.png} \hspace{-3em} \raisebox{12ex}{\includegraphics[width=0.5\textwidth]{museum2.png}}
        \caption{Interactive table with a plankton visualization. Image: \copyright{} 2016 IEEE, reprinted from Sauer et al.~\cite{Sauer2016}. TODO for layout: could you place the right image slightly overlapping on the left on? As to imply something similar to an inset?
  \label{fig:museum}}        
    \end{center}
\end{figure*}

@@

We want to share our perspective on what we see as important, ongoing trends. We adopt the point of view from visualization research as a whole, and then relate this to the particular interests for computing in science and engineering.

\subsection{The Interplay of Artificial Intelligence Techniques and Visualization}

Artificial Intelligence (AI) techniques play a crucial role in recognition tasks. In particular, machine learning (ML) algorithms are very efficient in pattern recognition problems such as image classification, speech recognition, and sentiment analysis. Over the past years, we have seen an increasing intersection between ML and visualization with the popularity of Deep Learning (DL). Unlike classical ML algorithms, DL algorithms process sets of labeled samples using multiple layers to extract features automatically.
A common intersection between AI and visualization is related to the field of explainable AI, where we use visualization techniques to gain insights into the inner workings of DL neural networks. We have recently seen several works that describe techniques to interpret DL models~\cite{hohman2018visual,GARCIA2018}. Most approaches aim at understanding how features are perceived during the learning process, in which visualization techniques can offer insights to improve the process.

The complexity of DL architectures also benefits from visualization to identify which aspects of the network can be modified. Finally, the understanding of the inner working of a deep model, a process called training analysis, has also been drawing attention recently by including the human in the loop. Such techniques are being referred to as interactive visual machine learning (IVML). As the popularity of DL grows steadily, we foresee that visualization techniques will continue to play a crucial role in understanding and improving new approaches.

With the increasing use of ML and especially DL in science and engineering, this trend is highly relevant for us as well.

\subsection{Getting Insights from Data: Visualization in Data Science}

Data is at the center of essential tasks in our society, such as decision-making, planning strategies, or process optimizations, and of course, also in data-driven science and engineering. The extraction of insights from data is the main focus of data science. There are, however, challenges in this process, since data is often massive, heterogeneous, subject to errors, and comes in a streaming fashion. Along with other fields such as applied mathematics, statistics, ML (see above), and computational modeling, visualization plays a fundamental part in data analysis. Urban data is a good example that can benefit from visualization techniques, such as spatio-temporal representations of data~\cite{Doraiswamy2018}. Several works describe ways to understand mobility patterns in cities from their fleet of taxis, trains, or bike-sharing systems. 
%Sports is another source of data that significantly benefit from visualization techniques. 
%Several approaches use visualization to understand soccer, tennis, and baseball matches, among other sports~\cite{perin2018}. 
Another example is social media. Visualization techniques are being used to display and explore social networks and their flow of information diffusion, or even explore the textual content shared among users~\cite{Chen2017}. Visual text analysis, which combines visualization along with other areas such as data mining, natural language processing, and machine learning, offers great ways to extract insights from social media. In all these examples we see how visualization techniques allow the extraction of insights from large, heterogeneous and unstructured data. In science and engineering we observe a growing need of this type of data and similar needs for insights.

@@

\subsection{Integration of VIS Subfields}

IEEE VIS has been structured around its three main conferences: the IEEE Conference on Visual Analytics Science and Technology (VAST), the IEEE Conference on Information Visualization (InfoVis), and the IEEE Conference on Scientific Visualization (SciVis). These conferences have traditionally targeted different subfields of visualization. For example, scientific visualization primarily considers data from scientific applications, i.e., it is the natural link to computing in science and engineering. Scientific visualization usually deals with data with a given spatial embedding such as 3D scalar or vector fields. In contrast, information visualization typically works data without a given spatial mapping; for example, the visualization of networks or high-dimensional data. Melanie Tory and Torsten M{\"o}ller provide more details on spatial and nonspatial visualization \cite{ToryM04}. Finally, visual analytics often includes ML techniques for improved data analysis and addresses the challenges of data science.

We have already seen in the first part of this article that the separation of the three subfields is not so clear in practice. For example, multidimensional or multivariate data is highly interesting in science and engineering and, therefore, appropriate visualizations will likely include both SciVis and InfoVis techniques. We have seen a lot of highly exciting papers that combine techniques from the subfields of VIS; these are particularly interesting for computing in science and engineering. 
IEEE VIS reflects this trend toward combined techniques by integrating the VAST, InfoVis, and SciVis Conferences, according to the results of a restructuring process (http://ieeevis.org/governance/restructuring).

Besides this integration, we have also been witnessing visualization as a growing and maturing field. Therefore, fundamental and long-standing research challenges are still being addressed, but now on a more refined and advanced level. As an example that is particularly relevant for computing in science and engineering, there is ongoing research on making visualization useful for applications in high-performance computing (HPC), by exploiting in-situ visualization integrated into the HPC environment or developing efficient, parallel visualization techniques. Another ongoing topic targets improved mathematical methods for visualization, for example, relying on topological techniques or uncertainty quantification.

@@

\subsection{Visualization and Human-Computer Interaction}

We have also been witnessing an increasingly strong link between visualization and human-computer interaction (HCI) research. One theme of our retrospective addresses HCI from the visualization perspective. With interactive data exploration, interaction techniques play a highly significant role in visualization, with a large number of papers at IEEE VIS. However, visualization has also become interesting for HCI research as a relevant topic. This is reflected by the fact that the leading HCI conference---ACM CHI (https://chi.acm.org)---included a separate subcommittee for visualization for CHI 2020. This link nicely strengthens the human-centered perspective on visualization.
Furthermore, it is the right way of addressing the long-standing and challenging issue of evaluating visualization approaches from the user's perspective. For example, the series of BELIV Workshops (https://beliv-workshop.github.io/) targets this issue. While these HCI and user-based evaluation problems are primarily interesting for basic visualization research, we expect that improved interactive visualization will eventually impact any visualization application, including ones in science and engineering.

@@

We think that our discussion of past VizCorner papers from the last ten years and the outlook on trends show that visualization is a thriving and evolving research field with many interesting connections to computing in science and engineering. With this paper, the tenure of one of the current co-editors (Daniel Weiskopf) will end. 
I, Daniel, have been serving in this role since 2011, and it has been a great pleasure working for the VizCorner. Thanks to all readers and authors of the VizCorner, and special thanks to my former co-editor Claudio Silva, my current co-editor Joao Comba, and the whole CiSE team!

The VizCorner with continue with a new incoming co-editor joining Joao. This change of co-editor will allow us to refresh the topics of the VizCorner. We look forward to bringing interesting new articles in the years to come.

>>>>

\begin{figure*}
    \begin{center}    
        \includegraphics[width=0.40\textwidth]{museum1.png} %\hspace{-3em} \raisebox{12ex}{\includegraphics[width=0.5\textwidth]{museum2.png}}
        \includegraphics[width=0.48\textwidth]{museum2.png}
        \caption{Interactive table with a plankton visualization. Image: \copyright{} 2016 IEEE, reprinted from Sauer et al.~\cite{Sauer2016}. TODO for layout: could you place the right image slightly overlapping on the left on? As to imply something similar to an inset?
  \label{fig:museum}}        
    \end{center}
\end{figure*}

@@

We want to share our perspective on what we see as important, ongoing trends. We adopt the point of view of visualization research as a whole, and address the particular interests for computing in science and engineering.

\subsection*{Interplay of artificial intelligence and visualization}

Artificial Intelligence (AI) techniques play a crucial role in recognition tasks. In particular, machine learning (ML) algorithms are efficient in pattern-recognition problems such as image classification, speech recognition, and sentiment analysis. Over the past years, we have seen an increasing intersection between ML and visualization with the popularity of Deep Learning (DL). Unlike classical ML algorithms, DL algorithms process sets of labeled samples using multiple layers to extract features automatically.
A common intersection between AI and visualization is related to the field of explainable AI, where we use visualization techniques to gain insights into the inner workings of DL neural networks. We have recently seen several works that describe techniques to interpret DL models~\cite{hohman2018visual,GARCIA2018}. Most approaches aim at understanding how features are perceived during the learning process, in which visualization techniques can offer insights to improve the process.

The complexity of DL architectures also benefits from visualization to identify which aspects of the network can be modified. Finally, the understanding of the inner working of a deep model, a process called training analysis, has also been drawing attention recently by including the human-in-the-loop. Such techniques are being referred to as interactive visual machine learning (IVML). As the popularity of DL grows steadily, we foresee that visualization techniques will continue to play a crucial role in understanding and improving new approaches.

\subsection*{Getting insights from data: visualization in data science}

Data is at the center of essential tasks in our society, such as decision-making, planning strategies, or process optimizations, and of course, also in data-driven science and engineering. 
Extracting insights from data is challenging because data is often massive, heterogeneous, subject to errors, and comes in a streaming fashion. Along with other fields such as applied mathematics, statistics, ML (see above), and computational modeling, visualization plays a fundamental part in data analysis. Urban data is a good example that can benefit from visualization techniques, such as spatio-temporal representations of data~\cite{Doraiswamy2018}. Several works describe ways to understand mobility patterns in cities from their fleet of taxis, trains, or bike-sharing systems. 
%Sports is another source of data that significantly benefit from visualization techniques. 
%Several approaches use visualization to understand soccer, tennis, and baseball matches, among other sports~\cite{perin2018}. 
Another example is social media. Visualization techniques are being used to display and explore social networks and their flow of information diffusion, or even explore the textual content shared among users~\cite{Chen2017}. Visual text analysis, which combines visualization along with other areas such as data mining, natural language processing, and machine learning, offers great ways to extract insights from social media. In all these examples we see how visualization techniques allow the extraction of insights from large, heterogeneous and unstructured data. In science and engineering we observe a growing need for this type of data and similar needs for insights.

@@

\subsection*{Integration of VIS subfields}

IEEE VIS has been structured around its three main conferences: the IEEE Conference on Visual Analytics Science and Technology (VAST), the IEEE Conference on Information Visualization (InfoVis), and the IEEE Conference on Scientific Visualization (SciVis). 
Scientific visualization primarily considers data from scientific applications, i.e., it is the natural link to computing in science and engineering. It usually deals with data with a given spatial embedding such as 3D scalar or vector fields. Information visualization typically works with data lacking a spatial mapping; for example, the visualization of networks or high-dimensional data. Melanie Tory and Torsten M{\"o}ller provide details on spatial and non-spatial visualization \cite{ToryM04}. Finally, visual analytics often includes ML techniques for improved data analysis and addresses the challenges of data science.

The separation of the three subfields is not so clear in practice. For example, multidimensional or multivariate data is highly interesting in science and engineering and, therefore, appropriate visualizations will likely include both SciVis and InfoVis techniques. We have seen a lot of exciting papers that combine techniques from the subfields of VIS; these are particularly interesting for computing in science and engineering. 
IEEE VIS reflects this trend toward combined techniques by integrating the VAST, InfoVis, and SciVis Conferences, according to the results of a restructuring process (http://ieeevis.org/governance/restructuring).

Besides this integration, we have also been witnessing visualization as a growing and maturing field. Fundamental and long-standing research challenges are still being addressed, but now on a more refined and advanced level. As an example that is particularly relevant for computing in science and engineering, there is ongoing research on making visualization useful for applications in high-performance computing (HPC), by exploiting in-situ visualization integrated into the HPC environment or developing efficient, parallel visualization techniques. Another ongoing topic targets improved mathematical methods for visualization, for example, relying on topological techniques or uncertainty quantification.

@@

\subsection*{Visualization and human-computer interaction}

We have also been witnessing an increasingly strong link between visualization and human-computer interaction (HCI) research. One theme of our retrospective addresses HCI from the visualization perspective. With interactive data exploration, interaction techniques play an important role in visualization, with a large number of papers at IEEE VIS. Visualization has also become interesting for HCI research. This is reflected by the fact that the leading HCI conference---ACM CHI (https://chi.acm.org)---included in 2020 a separate subcommittee for visualization. This link strengthens the human-centered perspective on visualization.
It is the right way of addressing the long-standing and challenging issue of evaluating visualization approaches from the user's perspective. For example, the series of BELIV Workshops (https://beliv-workshop.github.io/) targets this issue. While these HCI and user-based evaluation problems are primarily interesting for basic visualization research, we expect that improved interactive visualization will eventually impact any visualization application, including ones in science and engineering.

@@

Our retrospective of the VizCorner in the last ten years and the outlook on trends show that visualization is a thriving and evolving research field with many interesting connections to computing in science and engineering. With this paper, the tenure of one of the current co-editors (Daniel Weiskopf) will end. 
I, Daniel, have been serving in this role since 2011, and it has been a great pleasure working for the VizCorner. Thanks to all readers and authors of the VizCorner, and special thanks to my former co-editor Claudio Silva, my current co-editor Joao Comba, and the whole CiSE team!

The VizCorner will continue with a new incoming co-editor joining Joao, who will help us to refresh the topics covered in the department. We look forward to bringing interesting new articles in the years to come.
